SPARQL on the Open, Decentralised Web

### 2016-07-22 dissertation/index.tex
note draft
A start of a new intro
add mainmatter marker
More restructuring
Helen's corrections to intro

\begin{abstract}
  This dissertation discusses a broad range of problems concerning the
  use of the SPARQL query language on the open, public Web. It is
  motivated from seeing decentralisation of data and infrastructure as
  an important social goal, and SPARQL as an enabling technology to
  solve problems using the Semantic Web. The dissertation makes
  contributions in hypermedia, where RDF is used to create a format
  that can tell humans and machines alike how to manipulate resources
  on the Web; philosophy of science, where important foundational
  problems around how to create valid knowledge and objectivity is
  discussed; statistical methods that better satisfy the
  requirements from philosophy of science than current practice; how
  to improve developer efficiency with novel programming paradigms;
  and finally how caching infrastructure in the Internet may be used
  to make query answering across the Web more robust.
\end{abstract}

\tableofcontents

\chapter{Introduction}

\input{introdefhistory}
\input{motivation}
\input{related}
\input{problems}
\input{papersincontext}
\input{conclusions}
\input{cache-proxy}

>>>>

\begin{abstract}
  This dissertation discusses a broad range of problems concerning the
  use of the SPARQL query language on the open, public Web. It is
  motivated from seeing decentralisation of data and infrastructure as
  an important social goal, and SPARQL as an enabling technology to
  solve problems using the Semantic Web. The dissertation makes
  contributions in hypermedia, where RDF is used to create a format
  that can tell humans and machines alike how to manipulate resources
  on the Web; philosophy of science, where important foundational
  problems around how to create valid knowledge and objectivity are
  discussed; statistical methods that better satisfies the
  requirements from philosophy of science than current practice; how
  to improve developer efficiency with novel programming paradigms;
  and finally how caching infrastructure in the Internet may be used
  to make query answering across the Web more robust.

  \textbf{This is an unfinished draft of the thesis.}
\end{abstract}

\tableofcontents
\mainmatter

\input{introduction}

\input{introdefhistory}
\input{motivation}

\chapter{Technology Details}
\input{details}
\input{related}

\chapter{Problems and Solutions}

\input{problems}
\input{papersincontext}
\input{conclusions}
\input{cache-proxy}

### 2016-07-22 dissertation/introdefhistory.tex
Helen's corrections to intro

The Semantic Web can be described as a machine-readable Web of Data,
defined by a technology stack managed by the World Wide Web
Consortium. While individual technologies can be given precise,
operational definitions, the above definition is naive and can hardly
be used operationally to derive new knowledge. The Semantic Web is a
complex, human artefact that must be understood not only in technical
terms, but also in social, economic and cultural terms. First, the Web
is not only readable, but also writeable. Whether the essential parts
of a data integration problem will be solved by e.g. pervasive
ontology alignment, a linked open vocabularies approach where many
players adopt a fairly consistent set of vocabularies, or by major
players that are able to force convergence towards e.g. \texttt{Schema.org}
is an example of the social, economic and cultural mechanisms that
shape the future of the Semantic Web, see
Section~\ref{sec:semproblems} for a more elaborate
discussion. Likewise, it may not be the W3C Semantic Web stack at all
that achieves success, though unlikely at this point, it may be
something in the extension of microformats\footnote{See
  \url{http://microformats.org/}}, microdata\cite{microdata}, or
something else entirely. For a thoughtful criticism of Linked Data
with a proposed alternative, see \cite{darobin1}, but again, it is
likely that social, economic and cultural factors will be as decisive
as technical. One may say that e.g. microformats cannot help achieve
the visions that may be stated for the Semantic Web, but that again
emphasises visions, not operational definitions.

@@

The problems are also discussed throughout
\cite{chalmers1999whatis}. It is interesting to consider the example
on page 25, where the author describes a method forwarded by Galileo
Galilei to measure the diameter of a star. The method was objective in
the sense if followed, it would yield the same value today as it did
when Galileo employed it. However, it relied on faulty assumptions and
is therefore invalid. The same kind of problem may occur in the
evaluation of e.g. SPARQL engines. If a benchmark relies on a faulty
workload, the application of the benchmark may seem objective to the
investigator, but its external validity should obviously be
challenged. Moreover, if the benchmark relied on a use case supplied
by e.g. an industrial partner, the validity could be further
challenged based on e.g. the restriction on viewpoint that this
represents.

@@

I was convinced that the things that made the Web great to work with
had to be present in the Semantic Web as well: Everybody is empowered
to publish, find and quickly make use of Free Software, learn by
``View Source'' (an idea which becomes even more powerful with
semantics, as if you can understand what you are viewing simply by
looking at the message, you can begin processing it), anyone can mint
identifiers, anybody can reuse those identifiers in their own
contexts, and eventually graduate to reading technical materials as
needed. Many of these things are virtues of the Semantic Web as it is
an extension of the Web, others have not been sufficiently cultivated
or even appreciated. Moreover, my experience corresponds to the
virtues that Tim Berners-Lee has attributes the Web's success
to\footnote{See
  \url{http://dig.csail.mit.edu/2007/03/01-ushouse-future-of-the-web}}.

@@

Unfortunately, management did not permit more time to enhance and document the
SPARQL Endpoint further, and it didn't see much practical use. 
I was, however, allowed to join several World Wide Web Consortium
groups. First, the Web Content Label Incubator Group, which was tasked
with discussing a replacement for the then archaic Platform for
Internet Content Selection, and resolved to use RDF for this
purpose, and it transitioned into the POWDER Working Group, which
provided the specifications.

More importantly, I joined the Semantic Web Education and Outreach
(SWEO) Interest Group, where I admitted my frustration with the lack
of practical progress with Semantic Web technologies and the relative
lack of uptake. This sentiment was echoed by the late Aaron Swartz,
who was asked by the group on his opinion, and he emailed the
following statement\footnote{See
  \url{https://lists.w3.org/Archives/Public/public-sweo-ig/2006Dec/0138.html}}:
\begin{quote}
I'm not sure what SWEO is, but my feeling is and pretty much always
has been that the Semantic Web people need to start putting together
Genuinely Useful stuff that can be done Right Now.
\end{quote}

This was one of the motivations I had for starting the SWEO Community
Projects, where a questionnaire was posted on the Web, and people from
the Semantic Web community were challenged to come up with a project
that would provide practical benefits in the short term. We received
10 proposals, out of which 3 was selected for backing by the SWEO
IG. None of them were successful, but a fourth proposal, submitted by
Chris Bizer and Richard Cyganiak built momentum quickly and was
therefore also selected for backing, despite some criticism by several
IG members, myself included. The proposal was titled ``Linking Open
Data'' and sought to take already abundant open data and model it
using guidelines of the Linked Data Design
Issue\cite{linkeddataissue}. LODstats\cite{auer2012lodstats} provides
extensive statistics on the results of this project, and has of this
writing seen nearly 10~000 data sets.

At this point, I left Opera to work as a consultant on Semantic Web
technologies. One of the projects that were successfully completed was
called Sublima\cite{sublima}. It was driven entirely by RDF and SPARQL,
and featured faceted navigation, navigation in a thesaurus, and full
text search. Experiences from this project influenced requirements of
SPARQL 1.1 \cite{sparql11query}, and I was an editor of the SPARQL 1.1 Features and
Rationale specification\cite{sparql11new}. Parts of the property paths
feature and aggregate queries were used extensively. Write operations
were also implemented with SPARQL using what was then specified only
in a member submission to the W3C\cite{seaborne2008sparql}. This was the basis for
the update language in SPARQL 1.1 but differed substantially in
surface syntax, partly because our work showed how certain usage patterns
could be simplified.

Two features that arose from Sublima requirements were not accepted
for SPARQL 1.1, however: Full text index and a feature known as
``Limit Per Resource''\footnote{See
  \url{http://www.w3.org/2009/sparql/wiki/Feature:LimitPerResource}}. The
latter feature arises from the fact that usually, the user is
indifferent to the number of solutions to a query, and therefore
limiting by the number of solutions is unhelpful. This is especially
true since other features of SPARQL, such as optional clauses, are
helpful in dealing with heterogeneous data. In the Sublima case, it
was interesting to limit by the number of articles that were returned,
but the number of solutions depended on the number of authors of an
article and the concepts used to classify an article. To achieve the
desired effect with SPARQL 1.1, one would have to write a very complex
subquery, which is undesirable for such an important feature.

@@

However, it became clear that I could not have sufficient time in the
industry to pursue what was developing as my main interest, query
answering over data on the open Web for generic application
development.

>>>>

The Semantic Web can be described as a machine-readable Web of Data,
defined by a technology stack managed by the World Wide Web
Consortium. While individual technologies can be given precise,
operational definitions, the above definition is naive and can hardly
be used operationally to derive new knowledge. The Semantic Web is a
complex, human artefact that must be understood not only in technical
terms, but also in social, economic and cultural terms. First, the Web
is not only readable, but also writeable. Whether the essential parts
of a data integration problem will be solved by e.g. pervasive
ontology alignment, a linked open vocabularies approach where many
players adopt a fairly consistent set of vocabularies, or by major
players that are able to force convergence towards e.g. \texttt{Schema.org}
is an example of the social, economic and cultural mechanisms that
shape the future of the Semantic Web, see
Section~\ref{sec:semproblems} for a more elaborate
discussion. Likewise, it may not be the W3C Semantic Web stack at all
that achieves success, though unlikely at this point, it may be
something in the extension of microformats\footnote{See
  \url{http://microformats.org/}}, microdata\cite{microdata}, or
something else entirely. For a thoughtful criticism of Linked Data
with a proposed alternative, see \cite{darobin1}, but again, it is
likely that social, economic and cultural factors will be as decisive
as technical. One may say that, for example microformats cannot help achieve
the visions that may be stated for the Semantic Web, but that again
emphasises visions, not operational definitions.

@@

The problems are also discussed throughout
\cite{chalmers1999whatis}. It is interesting to consider the example
on page 25, where the author describes a method forwarded by Galileo
Galilei to measure the diameter of a star. The method was objective in
the sense if followed, it would yield the same value today as it did
when Galileo employed it. However, it relied on faulty assumptions and
is therefore invalid. The same kind of problem may occur in the
evaluation of e.g. SPARQL engines. If a benchmark relies on a faulty
workload, the application of the benchmark may seem objective to the
investigator, but its external validity should obviously be
challenged. Moreover, if the benchmark relied on a use case supplied
by for example an industrial partner, the validity could be further
challenged based on e.g. the restriction on viewpoint that this
represents.

@@

I was convinced that the things that made the Web great to work with
had to be present in the Semantic Web as well: Everybody is empowered
to publish, find and quickly make use of Free Software, learn by
``View Source'' (an idea which becomes even more powerful with
semantics, as if you can understand what you are viewing simply by
looking at the message, you can begin processing it), anyone can mint
identifiers and anybody can reuse those identifiers in their own
contexts, and eventually graduate to reading technical materials as
needed. Many of these things are virtues of the Semantic Web as it is
an extension of the Web, others have not been sufficiently cultivated
or even appreciated. Moreover, my experience corresponds with the
virtues that Tim Berners-Lee has attributed the Web's success\footnote{See
  \url{http://dig.csail.mit.edu/2007/03/01-ushouse-future-of-the-web}}.

@@

Unfortunately, management did not permit more time to enhance and document the
SPARQL Endpoint further, and it didn't see much practical use. 
I was, however, allowed to join several World Wide Web Consortium
groups. Firstly, the Web Content Label Incubator Group, which was
given the task of discussing a replacement for the then archaic Platform for
Internet Content Selection, and resolved to use RDF for this
purpose. It transitioned into the POWDER Working Group, which
provided the specifications.

Secondly and more importantly, I joined the Semantic Web Education and Outreach
(SWEO) Interest Group, where I admitted my frustration with the lack
of practical progress with Semantic Web technologies and the relative
lack of uptake. This sentiment was echoed by the late Aaron Swartz,
who when asked by the group on his opinion, emailed the
following statement:
\begin{quote}
I'm not sure what SWEO is, but my feeling is and pretty much always
has been that the Semantic Web people need to start putting together
Genuinely Useful stuff that can be done Right Now.\footnote{See
  \url{https://lists.w3.org/Archives/Public/public-sweo-ig/2006Dec/0138.html}}
\end{quote}

This was one of my motivations for starting the SWEO Community
Projects, where a questionnaire was posted on the Web, and people from
the Semantic Web community were challenged to come up with a project
that would provide practical benefits in the short term. We received
10 proposals, out of which 3 were selected for backing by the SWEO
IG. None of them were successful, but a fourth proposal, submitted by
Chris Bizer and Richard Cyganiak built momentum quickly and was
therefore also selected for backing, despite some criticism by several
IG members, myself included. The proposal was titled ``Linking Open
Data'' and sought to take already abundant open data and model it
using guidelines of the Linked Data Design
Issue\cite{linkeddataissue}. LODstats\cite{auer2012lodstats} provides
extensive statistics on the results of this project, and has by the
time of this writing seen nearly 10~000 data sets.

At this point, I left Opera to work as a consultant on Semantic Web
technologies. One of the projects that were successfully completed were
called Sublima\cite{sublima}. It was driven entirely by RDF and SPARQL,
and featured faceted navigation, navigation in a thesaurus, and full
text search. Experiences from this project influenced requirements of
SPARQL 1.1 \cite{sparql11query}, and I was an editor of the SPARQL 1.1 Features and
Rationale specification\cite{sparql11new}. Parts of the property paths
feature and aggregate queries were used extensively. Write operations
were also implemented with SPARQL using what was then specified only
in a member submission to the W3C\cite{seaborne2008sparql}. This was the basis for
the update language in SPARQL 1.1 but differed substantially in
surface syntax, partly because our work showed how certain usage patterns
could be simplified.

Two features that arose from Sublima requirements were not accepted
for SPARQL 1.1: Full text index and a feature known as
``Limit Per Resource''\footnote{See
  \url{http://www.w3.org/2009/sparql/wiki/Feature:LimitPerResource}}. The
latter feature arises from the fact that usually, the user is
indifferent to the number of solutions to a query, and therefore
limiting by the number of solutions is unhelpful. This is especially
true since other features of SPARQL, such as optional clauses, are
helpful in dealing with heterogeneous data. In the Sublima case, it
was interesting to limit by the number of articles that were returned,
but the number of solutions depended on the number of authors of an
article and the concepts used to classify an article. To achieve the
desired effect with SPARQL 1.1, one would have to write a very complex
subquery, which is undesirable for such an important feature.

@@

However, it became clear that I would not have sufficient time in the
industry to pursue what was developing as my main interest, query
answering over data on the open Web for generic application
development.

### 2016-07-22 dissertation/introduction.tex
intro prelim
Helen's corrections to intro

\section{Semantic Web Technology}

The World Wide Web, or just the Web for short, is a well-known global
information space invented by Tim Berners-Lee in 1989, further emerged
in the 1990-ties. It is characterised first and foremost by its
universality, anyone can set up a computer, connect it to the Internet and
start serving data or documents from it, and further adapt it to their purpose.

>>>>

\section{Semantic Web Technology}\label{sec:semwebtech}

The World Wide Web, or just the Web for short, is a well-known global
information space invented by Tim Berners-Lee in 1989, further emerged
in the 1990s. It is characterised first and foremost by its
universality, anyone can set up a computer, connect it to the Internet and
start serving data or documents from it, and further adapt it to their purpose.
